---
title: "Week One"
output:
  xaringan::moon_reader:
    css: ["fc", "fc-fonts", "reed.css", "default"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-forest-light
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, fig.align = "center")
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)
```

# Case Study: Sampling

---
# Case Study: Sampling

We can estimate the mean $\mu$ of a population of size $N$ in two ways:

--

1. Take a simple random sample of size $n$ and calculate $\bar{x}$.

--

2. Take a stratified sample of $n_1$ from the first strata and $n_2$ from the 
second strata, proportionally, then compute $\bar{x}$.

--

Which estimate (r.v.) will have "better" properties?

---
# Set up population

```{r}
set.seed(403)
```

--

```{r}
# generate population
N_1 <- 20000
N_2 <- 30000
g1 <- rnorm(N_1, mean = 1, sd = 1)
g2 <- rnorm(N_2, mean = 3, sd = 1)
pop <- c(g1, g2)
```

--

```{r}
# lay out parameters
N <- N_1 + N_2
l_1 <- N_1/N
l_2 <- N_2/N
mu_1 <- mean(g1)
sigsq_1 <- var(g1) * (N_1 - 1) / N_1
mu_2 <- mean(g2)
sigsq_2 <- var(g2) * (N_2 - 1) / N_2
```


---
# 1. One simple random sample 

--

```{r fig.height=3, fig.align="center"}
n <- 100
samp <- sample(pop, size = n)
df <- data.frame(x = samp)
library(tidyverse)
ggplot(df, aes(x = x)) +
  geom_dotplot(alpha = .5) +
  theme_bw() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```


---
# 1. One simple random sample, cont 

```{r fig.height = 3}
# Density plot
ggplot(df, aes(x = x)) +
  geom_density(fill = "steelblue", alpha = .5) +
  theme_bw() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```


---
# 1. Simulation: SRS

--

```{r}
it <- 5000
xbar_srs <- rep(NA, it)
for (i in 1:it) {
  samp <- sample(pop, size = n)
  xbar_srs[i] <- mean(samp)
}
```

--

```{r}
head(xbar_srs)
```


---
# 2. One stratified sample 

--

```{r}
n_1 <- n * l_1
n_2 <- n * l_2
samp_1 <- sample(g1, size = n_1)
samp_2 <- sample(g2, size = n_2)
df <- data.frame(x = c(samp_1, samp_2),
                 stratum = rep(c("1", "2"), 
                               times = c(n_1, n_2)))
```

--

```{r, fig.height=3, echo = FALSE}
ggplot(df, aes(x = x, fill = stratum)) +
  geom_dotplot(alpha = .5) +
  theme_bw() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```


---
# 2. Simulation: Stratified sampling

--

```{r}
xbar_strat <- rep(NA, it)
for (i in 1:it) {
  samp_1 <- sample(g1, size = n_1)
  samp_2 <- sample(g2, size = n_2)
  xbar_strat[i] <- mean(c(samp_1, samp_2))
}
```

```{r}
head(xbar_strat)
```


---
# Simulations compared 

```{r, fig.height=3}
df <- data.frame(xbar = c(xbar_srs, xbar_strat),
                 method = rep(c("srs", "strat"), times = c(it, it)))
ggplot(df, aes(x = xbar, fill = method)) +
  geom_density(alpha = .5) +
  theme_bw() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```


---
# Simulation vs Analytical

--

Through simulation, we have an empirical approximation to the variances of each estimate that we can compare to our analytical estimates.

--

```{r}
#SRS
var(xbar_srs)
(1/n) * (var(pop) * (N-1)/N)
# stratified
var(xbar_strat)
(1/n) * (l_1 * sigsq_1 + l_2 * sigsq_2)
```


---
# Simulation vs Analytical, cont.

--

We can more directly compare the analytical variances by writing the SRS estimate in terms of the strata variances and finding the difference in the variances of the two estimates.

$$
\textrm{difference:} \quad \frac{1}{n} \sum_{j = 1}^J \lambda_j \left( \mu_j - \mu \right)^2
$$

---
# Simulation-Empirical-Computational Approach

--

- Helps build intuition
- Checks analytical result
- Widely applicable

--

# Analytical Approach

--

- When it is exists, it is:
    + more general
    + often yields insight