---
title: "Week Four"
output:
  xaringan::moon_reader:
    css: ["fc", "fc-fonts", "reed.css", "default"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-forest-light
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, fig.align = "center", 
                      warning = FALSE, fig.height = 5.5)
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)
```

# Estimation Theories Compared

---
# Bayesian Estimation
--

## Benefits

- Can incorporate other information
- Intuitive interpretation
- Clearly articulated loss function

--

## Limitations

- Requires specification of prior (should check sensitivity)
- High dimensional priors are hard to think about
- Some posteriors don't have closed-forms (no prob!)


---
# Likelihood-based Estimation
--

## Benefits

- Intuitive interpretation (in a limited sense)
- Under certain circumstances, MLEs are asymptotically Normal
- Easy to find (derivatives or numerical optimization)

--

## Limitations

- MLEs don't always exist
- MLE not always unique
- Doesn't tell us anything about $P(\theta | x)$

---
# Method of Moments
--

## Benefits

- Often very easy to computer
- Can be used as a first guess for other methods
- High dim $\theta$? No prob!

--

## Limitations

- Doesn't work if the moments don't exist
- Often have higher MSE
- Can produce estimates outside parameter space

---
# Sampling from the Normal
--

```{r}
set.seed(49)
mu <- 4
sigma <- 1.5
n <- 5
x <- rnorm(n = n, mean = mu, sd = sigma)
```

--

```{r}
x
```

--

```{r echo = FALSE, fig.height=2.5}
ggplot(data.frame(x), aes(x)) +
  geom_dotplot() +
  theme_bw()
```


---
# Sampling Distribution of $\bar{X}$
--

```{r}
xbars <- rep(NA, 500)
for (i in 1:500) {
  x <- rnorm(n = n, mean = mu, sd = sigma)
  xbars[i] <- mean(x)
}
```

--

```{r echo = FALSE, fig.height=3}
p1 <- ggplot(data.frame(xbars), aes(xbars)) +
  geom_histogram(aes(y=..density..),
                 fill = "steelblue", color = "black") +
  theme_bw()  +
  theme(panel.border = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ylab("") +
  ylim(0, .7) +
  xlim(2, 6)
p1
```


---
# Sampling Distribution of $\bar{X}$

```{r}
xbars <- rep(NA, 500)
for (i in 1:500) {
  x <- rnorm(n = n, mean = mu, sd = sigma)
  xbars[i] <- mean(x)
}
```


```{r echo = FALSE, fig.height=3}
p1 + stat_function(fun = dnorm, 
                   args = list(mean = mu, sd = sigma/sqrt(n)),
                   lwd = 1.5)
```


```{r eval = FALSE}
p1 + stat_function(fun = dnorm, 
                   args = list(mean = mu, 
                               sd = sigma/sqrt(n)))
```

---
# Sampling Distribution of $\bar{X}$

```{r}
{{xbars <- rep(NA, 50000)}}
for (i in 1:50000) {
  x <- rnorm(n = n, mean = mu, sd = sigma)
  xbars[i] <- mean(x)
}
```


```{r echo = FALSE, fig.height=3}
ggplot(data.frame(xbars), aes(xbars)) +
  geom_histogram(aes(y=..density..),
                 fill = "steelblue", color = "black") +
  theme_bw()  +
  theme(panel.border = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ylab("") +
  stat_function(fun = dnorm, 
                   args = list(mean = mu, sd = sigma/sqrt(n)),
                   lwd = 1.5) +
  ylim(0, .7) +
  xlim(2, 6)
```


```{r eval = FALSE}
p1 + stat_function(fun = dnorm, 
                   args = list(mean = mu, 
                               sd = sigma/sqrt(n)))
```

---
# Sampling Distribution of $\hat{\sigma}^2$
--

```{r}
chis <- rep(NA, 500)
for (i in 1:500) {
  x <- rnorm(n = n, mean = mu, sd = sigma)
  chis[i] <- sum((x - mean(x))^2) / (sigma^2)
}
```

--

```{r echo = FALSE, fig.height=3}
p1 <- ggplot(data.frame(chis), aes(chis)) +
  geom_histogram(aes(y=..density..),
                 fill = "steelblue", color = "black") +
  theme_bw()  +
  theme(panel.border = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ylab("") 
p1
```

---
# Sampling Distribution of $\hat{\sigma}^2$

```{r}
chis <- rep(NA, 500)
for (i in 1:500) {
  x <- rnorm(n = n, mean = mu, sd = sigma)
  chis[i] <- sum((x - mean(x))^2) / (sigma^2)
}
```

```{r echo = FALSE, fig.height=3}
p1 + stat_function(fun = dchisq, 
                   args = list(df = n - 1),
                   lwd = 1.5)
```

```{r eval = FALSE}
p1 + stat_function(fun = dchisq, 
                   args = list(df = n - 1),
                   lwd = 1.5)
```

---
# Recap of Important results
--

Let $X_1, X_2, \ldots, X_n \sim N(\mu, \sigma^2)$.

--

1. $\bar{X} \sim N(\mu, \sigma^2/n)$.

--

2. $\frac{\sum (X_i - \bar{X})^2}{\sigma^2} \sim \chi^2_{n-1}$.

--

3. $\bar{X}$ and $\frac{1}{n}\sum (X_i - \bar{X})^2$ are independent (only true for Normal R.VS.)


---
# Independence of $\bar{X}$ and $\hat{\sigma}^2$
--

```{r}
xbars <- rep(NA, 500)
chis <- rep(NA, 500)
for (i in 1:500) {
  x <- rnorm(n = n, mean = mu, sd = sigma)
  xbars[i] <- mean(x)
  chis[i] <- sum((x - mean(x))^2) / (sigma^2)
}
```

--

```{r fig.height = 4, fig.width = 4, echo = FALSE}
ggplot(data.frame(x = xbars, y = chis), 
       aes(x = x, y = y)) +
  geom_point() +
  theme_bw() +
  labs(x = "xbar", y = "chis")
```


---
# Confidence Intervals

---
# Point Estimate for $\sigma^2$
--

```{r}
n <- 10
mu <- 10
sigmasq <- 4
x <- rnorm(n = n, mean = mu, sd = sqrt(sigmasq))
xbar <- mean(x)
sigsq_mle <- (1 / n) * sum((x - xbar)^2)
```

--

```{r}
sigsq_mle
```

---
# Interval Estimate for $\sigma^2$
--

```{r echo = FALSE, fig.height = 3.5}
ggplot(data.frame(x = c(0, 30)), aes(x = x)) +
  stat_function(fun = dchisq, args = list(df = n - 1)) +
  theme_bw() +
  theme(panel.border = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ylab("") +
  xlab("Chi-square, df = 9")
```

--

```{r}
alpha <- .05
q1 <- qchisq(alpha/2, df = n - 1)
q2 <- qchisq(1 - alpha/2, df = n - 1)
```

---
# Interval Estimate for $\sigma^2$

```{r echo = FALSE, fig.height = 3.5}
ggplot(data.frame(x = c(0, 30)), aes(x = x)) +
  stat_function(fun = dchisq, args = list(df = n - 1)) +
  theme_bw() +
  theme(panel.border = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ylab("") +
  xlab("Chi-square, df = 9") +
  geom_vline(xintercept = q1, lty = 2) +
  geom_vline(xintercept = q2, lty = 2)
```

```{r}
alpha <- .05
q1 <- qchisq(alpha/2, df = n - 1)
q2 <- qchisq(1 - alpha/2, df = n - 1)
```

```{r echo = FALSE}
q1
q2
```

---
# Interval Estimate for $\sigma^2$, cont.
--

```{r}
lower <- sum((x - xbar)^2)/q2
upper <- sum((x - xbar)^2)/q1
```

--

```{r echo = FALSE}
lower
upper
```

--

```{r eval = FALSE}
(upper - lower) / 2
```

--

```{r echo = FALSE}
(upper - lower) / 2
```

--

```{r eval = FALSE}
sigsq_mle
```

--

```{r echo = FALSE}
sigsq_mle
```
