---
title: "Rejection Sampling"
subtitle: "Lessons from Buffon's Needle"
author: "Math 392"
header-includes:
   - \usepackage{amsmath}
output:
  xaringan::moon_reader:
    css: ["fc", "fc-fonts", "reed.css", "default"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-forest-light
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---

```{r include = FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      fig.align = "center",
                      fig.height = 5.4)
library(tidyverse)
library(patchwork)

# set plot aesthetics
library(wesanderson)
pal <- wes_palette("Royal1")
col_target <- "darkgray"
lwd_target <- 1.5
lty_target <- 1
col_proposal <- "darkgray"
lwd_proposal <- 1.1
lty_proposal <- 2
col_theta <- "steelblue"
lwd_theta <- 1.1
lty_theta <- 3
col_theta_star <- "orchid"
lwd_theta_star <- 1.1
lty_theta_star <- 3
```

# Buffon's Needle (1733)

Dorp a needle of unit length on to a wood floor of infinite expanse with plane of unit length. What is the probability the needle crosses a line between planks?

<!-- Draw diagram of planks and needles. -->

---
# Parameterization

$D$: random distance from the needle midpoint to the line, $0 < D < \frac{1}{2}$

$\theta$: random angle between the needle and the line, $0 < \theta < \pi$.

<br>
<br>
<br>
<br>
<br>
<br>

<!-- Draw diagram with parameters added -->


*Observe*: the needle will cross when $D' > D$

$$\sin(\theta) = \frac{D'}{1/2}; \quad D' = \frac{\sin({\theta})}{2}$$

that is, when $\frac{\sin(\theta)}{2} > D$.


---
# Geometric probability

$$P(\textrm{needle crosses line}) = P\left(\frac{\sin(\theta)}{2} > D\right)$$
<!-- Draw sample space of D and theta with sin function -->

---
# Monte Carlo approximation

```{r}
theta <- runif(1e6, 0, pi)
D <- runif(1e6, 0, 1/2)
mean(sin(theta)/2 > D)
2/pi
```


---
# Generating from $f(\theta)$

Our original sample of $\theta$.

```{r echo = TRUE, fig.height = 3.7}
data.frame(theta) %>%
  ggplot(aes(x = theta)) +
  geom_density() +
  theme_bw()
```

---
# Generating from $f(\theta)$, cont.

The $\theta$ that meet our condition.

```{r echo = TRUE, fig.height = 3.7}
data.frame(theta) %>%
  filter(sin(theta)/2 > D) %>%
  ggplot(aes(x = theta)) +
  geom_density() +
  theme_bw()
```

---
# Monte Carlo approximation rephrased

1. Generate $\theta_i \sim Unif(0, \pi)$
2. Generate $D \sim Unif(0, 1/2)$
3. Retain $\theta_i$ if $D < f(\theta_i)$

Or:

1. Generate $\theta_i \sim Unif(0, \pi)$
2. Retain $\theta_i$ with probability $\frac{f(\theta_i)}{1/2}$


---
# Rejection sampling

<!-- Draw multimodal f, a non uniform q, and envelope M * q -->

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

Let $f(\theta)$ be the density we wish to sample from and $q(\theta)$ another density that we are able to sample from. Select an $M > 0$ such that $M q(\theta) \ge f(\theta)$ for all values of $\theta$.

1. Generate $\theta_i \sim q(\theta)$
2. Generate $U_i \sim Unif(0, 1)$
3. Retain $\theta_i$ if $U_i < \frac{f(\theta_i)}{M q(\theta_i)}$


---
# Practice: Sampling from the Gamma

In groups, use rejection sampling to approximate $\textrm{Gamma}(2, 3)$. Plot your approximation against the true density.

```{r echo = FALSE}
p1 <- ggplot(data.frame(x = c(0, 4)), aes(x = x)) +
  stat_function(fun = dgamma, 
                args = list(2, 3),
                col = col_target,
                lwd = lwd_target,
                lty = lty_target,
                n = 300) +
  geom_area(stat = "function",
            fun = dgamma,
            args = list(2, 3),
            fill = col_target,
            lwd = 1.5,
            alpha = .5) +
  ylim(c(0, 1.4)) +
  theme_bw() +
  labs(x = expression(theta),
       y = "") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
p1
```


```{r eval = FALSE, echo = FALSE}
# Uniform q
theta_proposed <- runif(1e6, 0, 5)
f <- dgamma(theta_proposed, 2, 3)
q <- dunif(theta_proposed, 0, 5)
M <- max(f/q)
U <- runif(1e6)

theta_accepted <- theta_proposed[U < f / (M * q)]

length(theta_accepted)/1e6

Mdunif <- function(x, min, max) {
  M * dunif(x, min, max)
}

ggplot(data.frame(theta_accepted), aes(x = theta_accepted)) +
  geom_density() +
  stat_function(fun = Mdunif, args = list(0, 5)) +
  theme_bw()


# Exponential q
theta_proposed <- rexp(1e6, 1)
q <- dexp(theta_proposed, .1)
M <- max(f/q)
U <- runif(1e6)

theta_accepted <- theta_proposed[U < f / (M * q)]

length(theta_accepted)/1e6

Mdexp <- function(x, rate) {
  M * dexp(x, rate)
}

ggplot(data.frame(theta_accepted), aes(x = theta_accepted)) +
  geom_density() +
  stat_function(fun = Mdexp, args = list(.1)) +
  theme_bw()
```

